DSCI 445 - Algal Blooms

### 2020-12-14 paper.Rmd
Added changes to explanation of models and merge threshold.

Tried 5 variables
+/- day merge
Tried day threshold

The in situ data was not collected as frequently as the satellite imagery data.
To remedy this, an experimental approach was taken in which the FIXME data
was shifted by a certain number of days forward and backward. Intuitively, algae
blooms do not drastically alter the characteristics of a lake between consecutive days.
Interpolating surrounding observations with existing data allowed for many more
of the satellite observations to be used, even though there was no in situ data
accompanying the observations.

However, the aforementioned Interpolating approach certainly fails after a
certain range of days. The optimal shift was determined experimentally: Each
model was fit using data shifted from one to ten days in either direction, and
the cross-validation error was used to assess model accuracy. The specifics of
the experimental structure are described below. The model accuracy provided a
picture of the quality of interpolation provided by each shift. If data was
shifted too many days, the prediction accuracy of the algorithm should
deteriorate.

@@

The algae bloom data is spatiotemporal; the blooms are recorded in a certain
geographic location at a certain time. This makes the class of linear models
unsuitable for this problem. Linear models make assumptions about the structure
of the errors in the data generating process that are violated by spatiotemporal
data. Spatiotemporal data has errors which are correlated; as observations
closer in space or time will covary (CITE).

To work around the issue of violated assumptions in classical statistical
models, we opted to use three variants of non-parametric, tree based regression
models. The models used were Random Forest, Gradient Boosting and XGBoost. These
models do not make assumptions about the structure of errors and correlations in
the data, meaning spatiotemporal data is usable.

TODO Describe each model?

@@

Train/test split? ($\frac{1}{3}$ was held out)
Experimental variables were compared using 5-fold cross validation

>>>>

Satellite image data and in situ data were not collected on similar schedules.
This meant that many observations were lost from both datasets on an inner join.
To remedy this problem, a procedure was used referred to as a "merge threshold."
Using a merge threshold, the in situ data was shifted forward and backward $0,
1, \cdots, n$ days. For example, on a merge threshold of two, the in situ data
from any single observation on day $k$ would be used on day $k - 2, k - 1, k, k
+ 1$, and $k + 2$. In other words, the data from any single in situ observation
  would be used for the surrounding four days. The merge threshold procedure was
  done up to ten days of threshold.

However, the aforementioned approach certainly fails after a
certain range of days. The optimal shift was determined experimentally: Each
model was fit using data shifted from one to ten days in either direction, and
the cross-validation error was used to assess model accuracy. The specifics of
the experimental structure are described below. The model accuracy provided a
picture of the quality of interpolation provided by each shift. If data was
shifted too many days, the prediction accuracy of the algorithm should
deteriorate.

@@

The algae bloom data is spatiotemporal; the blooms are recorded in a certain
geographic location at a certain time. This makes the class of linear models
unsuitable for this problem. Linear models make assumptions about the structure
of the errors in the data generating process that are violated by spatiotemporal
data. Spatiotemporal data has errors which are correlated; as observations
closer in space or time will covary.

The models selected were tree-based: Random Forest, Boosted Trees, and XGBoost. Although
the classical formulation of Random Forests and Boosted Trees rely on the
assumption of independent and identically distributed observations, tree-based
models were chosen in hopes that the extra flexibility in these models would
help account for the temporal structure in the data. Unfortunately, there was no
explicit spatial variable that could be included in the data to account for
spatial dependence.

The Random Forest and Bosted Tree implementation from `scikit-learn` were used.
The XGBoost implementation was taken from the Python package `xgboost`.

@@

The chosen models were fit on a grid of parameters. The parameters altered were
merge threshold, variable subset, and type of model. That is, we fit every
possible combination of merge threshold from $0, \cdots, 10$, the selected
variable subsets, and the selected model types.

Within each of the aforementioned combinations, a training set and test set were
created from the merged data, with the training set compromising two-thirds of
the dataset, and the test set compromising one-third of the dataset.

After a training and test set were created, $5$-fold cross-validation was
performed. Models were compared via $5$-fold cross validation error on the
training dataset. The final model was selected via this error, and then results
were repoted on the test data set.
